A tool for designing digital filters for the Hankel and Fourier transforms in potential, diffusive, and wavefield modeling

### 2018-06-19 ManuscriptGeophysics2018.tex
Manuscript re-submission

% \documentclass[manuscript]{geophysics}
% \documentclass[paper,twocolumn,twoside]{geophysics}
\documentclass[paper]{geophysics}
% \documentclass[paper, revised]{geophysics}
% \documentclass[manuscript,revised]{geophysics}

@@

\address{
% \footnotemark[1]TU Delft,
% Building 23,
% Stevinweg 1 / PO-box 5048,
% 2628 CN Delft,
% E-mail: \href{mailto:E.C.Slob@tudelft.nl}{E.C.Slob@tudelft.nl};
% \footnotemark[2]Instituto Mexicano del Petróleo,
% Eje Central Lázaro Cárdenas Norte 152,
% Col. San Bartolo Atepehuacan C.P. 07730,
% Ciudad de México, México,
% E-mail: \href{mailto:dieter@werthmuller.org}{Dieter@Werthmuller.org};
% \footnotemark[3]Lamont-Doherty Earth Observatory, Columbia University,
% 305C Oceanography,
% 61 Route 9W, PO Box 1000,
% Palisades NY 10964-8000 US,
% E-mail: \href{mailto:KKey@ldeo.columbia.edu}{KKey@ldeo.columbia.edu}.
}

\author{%
% Dieter Werthmüller\footnotemark[1]\footnotemark[2], %
% Kerry Key\footnotemark[3], and %
% Evert C.\ Slob\footnotemark[1]%
}

\footer{}
\lefthead{}%Werthmüller et al.}
\righthead{Digital filter designing tool}

@@

By substituting $r = e^x$ and $l = e^{-y}$ we get
%
\begin{equation}
  e^x F(e^x) = \int^\infty_{-\infty} f(e^{-y})K(e^{x-y})e^{x-y}\mr{d}y\ .
  \label{eq:filtint}
\end{equation}
%
This can be re-written as a convolution integral and be approximated for an
$N$-point filter by
%
\begin{equation}
  F(r) \approx \sum^N_{n=1} \frac{f(b_n/r) h_n}{r}\ ,
  \label{eq:filtapprox}
\end{equation}
%
where $h$ is the digital linear filter, and the logarithmically spaced filter
abscissae is a function of the spacing $\Delta$ and the shift $\delta$,
%
\begin{equation}
  b_n = \exp\left\{\Delta(-\lfloor{(N+1)/2\rfloor}+n) + \delta\right\} \ .
  \label{eq:base}
\end{equation}
%
\new{Note to reviewers: In equation~\ref{eq:base} $N/2$ changed to
$\lfloor{(N+1)/2\rfloor}$.} From equation~\ref{eq:filtapprox} it can be
seen that the filter method requires $N$ evaluations at each $r$. For example,
to calculate the frequency domain result for 100 offsets with a 201\,pt filter
requires 20,100 evaluations in the wavenumber domain. This is why the DLF often
uses interpolation to minimize the required evaluations, either for $F(r)$ in
what is referred to as \emph{lagged convolution DLF}, or for $f(l)$, which we
here call \emph{splined DLF}.

@@

The basic steps are as follows:
\begin{enumerate}
  \item Evaluate RHS of check-function $f_\mr{C}$:\newline
    \old{$d_\mr{R} = f_\mr{C}.\mr{rhs}(r)$}\\
    \new{$d_\mr{R} = f_\mr{C}^\mr{rhs}(r)$}
  \item Loop over each value $\Delta_i$, $\delta_j$ (brute force\new{;
      parameters on the left-hand side are overwritten where $i$, $j$
      are not mentioned}):
    \begin{enumerate}
      \item Calculate filter base:\newline
        \old{$b_n = \exp\left\{\Delta_i(-N/2+n) + \delta_j\right\}$}\\
        \new{$b_n = \exp\left\{\Delta_i(-\lfloor{(N+1)/2\rfloor}+n) +
      \delta_j\right\}$}
      \item Get required RHS and LHS evaluation points:\newline
        $r_\mr{I} = p(b, N)$\newline
        $l_\mr{I} = b/r_\mr{I}$
      \item Invert for filter coefficients:\newline
        \old{$J_{ij} = \mr{solve}(f_\mr{I}.\mr{lhs}(l_\mr{I}),
        f_\mr{I}.\mr{rhs}(r_\mr{I}))$}\\
        \new{$J = \mr{solve}\left[f_\mr{I}^\mr{lhs}(l_\mr{I}),
        f_\mr{I}^\mr{rhs}(r_\mr{I})\right]$}
    \item Calculate numerically RHS of check-function $f_\mr{C}$ with current
      filter \old{$J_{ij}$}\new{$J$}:\newline
      \old{$d_\mr{F} = f_\mr{C}.\mr{lhs}(b/r)\cdot J_{ij}\ /\ r$}\\
      \new{$d_\mr{F} = f_\mr{C}^\mr{lhs}(b/r)\cdot J\ /\ r$}
      \item Store minimum recovered amplitude or maximum $r$ where relative
        error is less than the provided error:\newline
        $\chi_{ij} = g\left[
        \mr{argmin}\left(|(d_\mr{F}-d_\mr{R})/d_\mr{R}| >
        \mr{error}\right) -1\right]\ $, where $g$ is either $d_\mr{R}$
        or $1/r$.
    \end{enumerate}
  \item \old{Return filter}\new{Re-calculate (Steps 2a-2c) filter base and}
    coefficients which yield minimum amplitude or maximum $r$ (a local
    minimization can be run to polish the brute-force result):\newline
    \new{$i_\mr{best}$, $j_\mr{best}$ = $\mr{argmin}(\chi)$}\newline
    \new{recalculate $b$, $J$ with $i_\mr{best}$, $j_\mr{best}$}\newline
    \old{return $J\left[\mr{argmin}(\chi)\right] $}\newline
  \item \new{Return filter base and coefficients and error matrix (for visual
    QC):}\newline
    \new{return $b$, $J$, $\chi$}\newline
\end{enumerate}

@@

\subsection{Design}
\old{Figure~\ref{fig:201b} shows the solution spaces of four consecutive
inversion runs, where each run is a focus on a subsection of the previous run,
indicated by the red square. The filter length in this case is $N=201$, and the
theoretical transform pairs are as given, e.g., in
\cite{USGS.75.Anderson}:}
\new{Figures~\ref{fig:201a} and~\ref{fig:201b} show the solution spaces of
seven consecutive inversion runs using the following theoretical transform
pairs \citep[e.g.,][]{USGS.75.Anderson},}
%
\begin{align}
  \int^\infty_0\,l \exp\left(-al^2\right) J_0(rl)\,dl &=
  \frac{\exp\left(\frac{-r^2}{4a}\right)}{2a}\ , \\
  \int^\infty_0\,l^2 \exp\left(-al^2\right) J_1(rl)\,dl &=
  \frac{r}{4a^2} \exp\left(-\frac{r^2}{4a}\right)\ ,
  \label{eq:j01}
\end{align}
%
where $a$ was set to 5. In the algorithm you can provide one pair
for the inversion, and a different pair to get the minimum amplitude or the
maximum $r$. In this case the same transform pair was used for both the
inversion and the check of quality. The acceptable relative error was set to
1\,\%. The RHS evaluation parameter for the inversion was set so  that an
over-determined system was evaluated with $M = 2N$ equations, where $r_\mr{I}$
was logarithmically spaced from $\log_{10}[1/\max(b)] - 1$ to
$\log_{10}[1/\min(b)] + 1$, where $b$ are the filter abscissae as given in
equation~\ref{eq:base}. This is accomplished in \texttt{fdesign} by setting
parameter $r_\mr{def} = (1, 1, 2)$.

@@

Designing filters is to a large extent trial and error. All input variables
influence the outcome, and often a \old{good}\new{even better} filter is found
by sheer luck of choosing the right starting parameters\new{, as can be seen in
Figure~\ref{fig:201c} (h). However, once you are in this zone every filter is a
very good filter}. Each of the input parameters has an effect on the outcome.
It also depends heavily on the transform pairs $f_\mr{I;C}$; functions that
decay rapidly are generally better, as noted by earlier authors
\citep[e.g.][]{USGS.75.Anderson}. It also depends on the filter length, and
obviously heavily on the spacing and shift values. Another important point is
how you define the right-hand evaluation points of the inversion
($r_\mr{def}$). Evaluating corresponding transform pairs separately or jointly
also leads to different filter coefficients ($J_0$, $J_1$, or $J_0$ \& $J_1$,
or equally sine, cosine, or sine \& cosine); also if the real or the imaginary
part is used when complex transform pairs or \texttt{empymod} is used; and if
it is inverted for the minimum amplitude or for the maximum $r$.

@@

The presented, free and open-source code \texttt{fdesign} can be used to design
digital linear filters for the Hankel and Fourier transforms (and more
generally for any linear transform) using either analytical transform pairs or
1D subsurface models together with the EM-modeler \texttt{empymod}. The code is
available from GitHub \old{in the add-on repository}\new{as part} of
\texttt{empymod}\old{, called \texttt{empyscripts},} from version
\old{0.2.0}\new{1.7.0} onwards.

@@

% D.W. would like to thank the electromagnetic research group of Aleksandr
% Mousatov at the IMP for fruitful discussions, and Jürg Hunziker for valuable
% feedback which improved this manuscript considerably.
We thank the reviewers for valuable inputs not only regarding the manuscript,
but also regarding the code and its documentation.

>>>>

% \documentclass[manuscript]{geophysics}
\documentclass[paper,twocolumn,twoside]{geophysics}
% \documentclass[paper]{geophysics}
% \documentclass[paper, revised]{geophysics}
% \documentclass[manuscript,revised]{geophysics}

@@

\address{
\footnotemark[1]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft,
E-mail: \href{mailto:dieter@werthmuller.org}{Dieter@Werthmuller.org};
\href{mailto:E.C.Slob@tudelft.nl}{E.C.Slob@tudelft.nl};
\footnotemark[2]Instituto Mexicano del Petróleo,
Eje Central Lázaro Cárdenas Norte 152,
Col. San Bartolo Atepehuacan C.P. 07730,
Ciudad de México, México,
\footnotemark[3]Lamont-Doherty Earth Observatory, Columbia University,
305C Oceanography,
61 Route 9W, PO Box 1000,
Palisades NY 10964-8000 US,
E-mail: \href{mailto:KKey@ldeo.columbia.edu}{KKey@ldeo.columbia.edu}.
}

\author{%
Dieter Werthmüller\footnotemark[1]\footnotemark[2], %
Kerry Key\footnotemark[3], and %
Evert C.\ Slob\footnotemark[1]%
}

\footer{}
\lefthead{Werthmüller et al.}
\righthead{Digital filter designing tool}

@@

By substituting $r = e^x$ and $l = e^{-y}$ we get
%
\begin{equation}
  e^x F(e^x) = \int^\infty_{-\infty} f(e^{-y})K(e^{x-y})e^{x-y}\mr{d}y\ .
  \label{eq:filtint}
\end{equation}
%
This can be re-written as a convolution integral and be approximated for an
$N$-point filter by
%
\begin{equation}
  F(r) \approx \sum^N_{n=1} \frac{f(b_n/r) h_n}{r}\ ,
  \label{eq:filtapprox}
\end{equation}
%
where $h$ is the digital linear filter, and the logarithmically spaced filter
abscissae is a function of the spacing $\Delta$ and the shift $\delta$,
%
\begin{equation}
  b_n = \exp\left\{\Delta(-\lfloor{(N+1)/2\rfloor}+n) + \delta\right\} \ .
  \label{eq:base}
\end{equation}
%
From equation~\ref{eq:filtapprox} it can be seen that the filter method
requires $N$ evaluations at each $r$. For example, to calculate the frequency
domain result for 100 offsets with a 201\,pt filter requires 20,100 evaluations
in the wavenumber domain. This is why the DLF often uses interpolation to
minimize the required evaluations, either for $F(r)$ in what is referred to as
\emph{lagged convolution DLF}, or for $f(l)$, which we here call \emph{splined
DLF}.

@@

The basic steps are as follows:
\begin{enumerate}
  \item Evaluate RHS of check-function $f_\mr{C}$:\newline
    \old{$d_\mr{R} = f_\mr{C}.\mr{rhs}(r)$}\\
    \new{$d_\mr{R} = f_\mr{C}^\mr{rhs}(r)$}
  \item Loop over each value $\Delta_i$, $\delta_j$ (brute force\new{;
      parameters on the left-hand side are overwritten where $i$, $j$
      are not mentioned}):
    \begin{enumerate}
      \item Calculate filter base:\newline
        \old{$b_n = \exp\left\{\Delta_i(-N/2+n) + \delta_j\right\}$}\\
        \new{$b_n = \exp\left\{\Delta_i(-\lfloor{(N+1)/2\rfloor}+n) +
      \delta_j\right\}$}
      \item Get required RHS and LHS evaluation points:\newline
        $r_\mr{I} = p(b, N)$\newline
        $l_\mr{I} = b/r_\mr{I}$
      \item Invert for filter coefficients:\newline
        \old{$J_{ij} = \mr{solve}(f_\mr{I}.\mr{lhs}(l_\mr{I}),
        f_\mr{I}.\mr{rhs}(r_\mr{I}))$}\\
        \new{$J = \mr{solve}\left[f_\mr{I}^\mr{lhs}(l_\mr{I}),
        f_\mr{I}^\mr{rhs}(r_\mr{I})\right]$}
    \item Calculate numerically RHS of check-function $f_\mr{C}$ with current
      filter \old{$J_{ij}$}\new{$J$}:\newline
      \old{$d_\mr{F} = f_\mr{C}.\mr{lhs}(b/r)\cdot J_{ij}\ /\ r$}\\
      \new{$d_\mr{F} = f_\mr{C}^\mr{lhs}(b/r)\cdot J\ /\ r$}
      \item Store minimum recovered amplitude or maximum $r$ where relative
        error is less than the provided error:\newline
        $\chi_{ij} = g\left[
        \mr{argmin}\left(|(d_\mr{F}-d_\mr{R})/d_\mr{R}| >
        \mr{error}\right) -1\right]\ $, where $g$ is either $d_\mr{R}$
        or $1/r$.
    \end{enumerate}
  \item \old{Return filter}\new{Re-calculate (steps 2a-2c) filter base and}
    coefficients which yield minimum amplitude or maximum $r$ (a local
    minimization can be run to polish the brute-force result):\newline
    \new{$i_\mr{best}$, $j_\mr{best}$ = $\mr{argmin}(\chi)$}\newline
    \new{recalculate $b$, $J$ with $i_\mr{best}$, $j_\mr{best}$}\newline
    \old{return $J\left[\mr{argmin}(\chi)\right] $}\newline
  \item \new{Return filter base and coefficients and error matrix (for visual
    QC):}\newline
    \new{return $b$, $J$, $\chi$}\newline
\end{enumerate}

@@

\subsection{Design}
\old{Figure~\ref{fig:201b} shows the solution spaces of four consecutive
inversion runs, where each run is a focus on a subsection of the previous run,
indicated by the red square. The filter length in this case is $N=201$, and the
theoretical transform pairs are as given, e.g., in
}\cite{USGS.75.Anderson}\old{:}
\new{Figures~\ref{fig:201a} and~\ref{fig:201b} show the solution spaces of
seven consecutive inversion runs using the following theoretical transform
pairs} \citep[e.g.,][]{USGS.75.Anderson}\new{,}
%
\begin{align}
  \int^\infty_0\,l \exp\left(-al^2\right) J_0(rl)\,dl &=
  \frac{\exp\left(\frac{-r^2}{4a}\right)}{2a}\ , \\
  \int^\infty_0\,l^2 \exp\left(-al^2\right) J_1(rl)\,dl &=
  \frac{r}{4a^2} \exp\left(-\frac{r^2}{4a}\right)\ ,
  \label{eq:j01}
\end{align}
%
where $a$ was set to 5. In the algorithm you can provide one pair
for the inversion, and a different pair to get the minimum amplitude or the
maximum $r$. In this case the same transform pair was used for both the
inversion and the check of quality. The acceptable relative error was set to
1\,\%. The RHS evaluation parameter for the inversion was set so  that an
over-determined system was evaluated with $M = 2N$ equations, where $r_\mr{I}$
was logarithmically spaced from $\log_{10}[1/\max(b)] - 1$ to
$\log_{10}[1/\min(b)] + 1$, where $b$ are the filter abscissae as given in
equation~\ref{eq:base}. This is accomplished in \texttt{fdesign} by setting
parameter $r_\mr{def} = (1, 1, 2)$.

@@

Designing filters is to a large extent trial and error. All input variables
influence the outcome, and often \old{a good}\new{an even better} filter is
found by sheer luck of choosing the right starting parameters\new{, as can be
seen in Figure~\ref{fig:201c} (h). However, once you are in this zone every
filter is a very good filter}. Each of the input parameters has an effect on
the outcome. It also depends heavily on the transform pairs $f_\mr{I;C}$;
functions that decay rapidly are generally better, as noted by earlier authors
\citep[e.g.][]{USGS.75.Anderson}. It also depends on the filter length, and
obviously heavily on the spacing and shift values. Another important point is
how you define the right-hand evaluation points of the inversion
($r_\mr{def}$). Evaluating corresponding transform pairs separately or jointly
also leads to different filter coefficients ($J_0$, $J_1$, or $J_0$ \& $J_1$,
or equally sine, cosine, or sine \& cosine); also if the real or the imaginary
part is used when complex transform pairs or \texttt{empymod} is used; and if
it is inverted for the minimum amplitude or for the maximum $r$.

@@

The presented, free and open-source code \texttt{fdesign} can be used to design
digital linear filters for the Hankel and Fourier transforms (and more
generally for any linear transform) using either analytical transform pairs or
1D subsurface models together with the EM-modeler \texttt{empymod}. The code is
available from GitHub \old{in the add-on repository}\new{as part} of
\texttt{empymod}\old{, called \texttt{empyscripts},}\new{; the version
  corresponding to this publication is \texttt{v1.7.1.}}\old{ from version
  0.2.0 onwards}.

@@

D.W. would like to thank the electromagnetic research group of Aleksandr
Mousatov at the IMP for fruitful discussions, and Jürg Hunziker for valuable
feedback which improved this manuscript considerably. We thank the reviewers
for valuable inputs not only regarding the manuscript, but also regarding the
code and its documentation.
